Paleoenvironmental models for Australia and the impact of aridification on blindsnake diversification

### 2023-04-17 manuscript.Rmd
changes made in response to suggestions from reviewers round 1

**Running title** blindsnake speciation slows down with aridification

@@

# Acknowledgements
ST would like to thank J Rolland and F. Condamine for helpful advice about fitting diversification models with RPANDA. Thanks to NJ Matzke and O Jiménez-Robles for help with BioGeoBEARS. Special thanks to IG Brennan, D Esquerré, M Pepper, CJ Pavón-Vázquez, and O Hagen for helpful advice on the analysis and discussion.

@@

**Aim**: Shifts in diversification rates of Australian flora and fauna have been associated with aridification, but the relationship between diversification rates and aridity has never been quantified. We employed multiple approaches to accurately reconstruct paleoaridity of Australia for the first time. We  used this information and phylogenetic-based analyses to explore how changes in temperature and increasing aridity during the Neogene influenced the diversification of the Australian blindsnakes and we further test whether diversification rates differ between arid-adapted and mesic-adapted lineages.

**Taxon**: blindsnakes

@@

**Results**: Ancestral area estimation indicated Australian blindsnakes have tropical grassland origins. We found that Australia-specific regional paleotemperature and paleoaridity provided a better explanation for diversification rate variation than global paleotemperature. Specifically, our best-fitting model indicate that speciation rates of blindsnakes decreased with increasing aridity. We found no difference in diversification rates between arid- and mesic-adapted lineages.

**Main conclusions**: Soon after colonising Australia, the common ancestors of Australian blindsnakes diversified rapidly in mesic habitats during the early Miocene. However, as the continent became increasingly arid, diversification rates decreased. We found that shifts in the environment led to the emergence of two major clades: one remaining in primarily mesic habitats and the other adapting to the expanding arid biome. Our results emphasise the importance of both arid and tropical biomes as sources and sinks of diversification.

@@

The present-day distribution of extant species can provide clues about their biogeographic history and insight into the tempo of their diversification. In particular, one major aim of evolutionary biology is to understand the role that environmental factors have played in shaping rates of species diversification. A variety of environmental correlates have been associated with diversification processes across clades, including elevation [@esquerreHow2019;@boschmanMountain2022], CO<sub>2</sub> concentration [@kergoatOpposite2018], microhabitats [@bars-closelDiversification2017], temperature [@ricklefsGlobal2006;@mayhewBiodiversity2012;@condamineMacroevolutionary2013;@condamineClimate2019;@rollandContribution2019;@sunRecent2020], and aridification [@raboskyExceptional2007;@pepperGenetic2011;@deepakAridification2018;@catulloAridification2014]. However, despite advancements in our understanding of the evolutionary relationships and divergence times of many clades, our knowledge of the impact of paleoenvironmental change on these clades remains limited, particularly in regard to the regional scale at which many clades radiate [@eyresClimatic2021]. As such, it remains a key challenge to combine regional paleoenvironmental histories, with molecular phylogenies, to understand the drivers of diversification over deep-time scales.

Terrestrial ectotherms, such as squamate reptiles, are highly sensitive to changes in external environment and climate. This can influence their diversification rates through multiple mechanisms. For instance, warmer temperature may increase resource availability, which in turn increase ecological niche space [@clarkeClimate2006]. Spatial and temporal heterogeneity in temperature can lead to climatic niche divergence, such as local adaptations in different geographic areas (Pepper et al., 2013), and change in activity time of prey/predators [@santosTemperature2011;@lawsonLatitudinal2014]. Change in climate, such as glaciation or aridification, has been shown to cause fragmentation of continuous habitats leading to isolated populations, which may lead to divergence [@oliverRocks2019]. Ectotherms are subjected to higher metabolic rates in warm climates (i.e. Kleiber’s law), which may increase mutation rates and ultimately greater speciation rates [@allenKinetic2006;@makarievaMean2008]. Finally, aridification may offer new ecological niche for species to exploit, particularly mesic-adapted lineages that have colonised from elsewhere.

The Australian continent has one of the most diverse reptile faunas on the planet and this may have been driven by its rapid environmental change in the Neogene period. After a long period of relatively stable climate and mesic habitats, the aridification of continental Australia began around 20 Ma (mega annum) with the development of pockets of aridity [@martinCenozoic2006;@byrneDecline2011]. Subsequent extreme warming and cooling periods in the last ~5 Ma creating what is now known as the "outback" [@byrneBirth2008;@pepperLife2021]. During this period, as aridification progressed, arid-adapted vegetation expanded while mesic biomes became increasingly fragmented [@crispHow2013;@martinCenozoic2006]. This shift in the environment also led to a radiation of some reptile clades, such as elapid snakes and skinks, which diversified into morphologically and ecologically distinct lineages [@leeDiversification2016;@sandersMolecular2008;@raboskyExceptional2007]. Conversely, other reptile clades, such as blindsnakes and geckos, remain morphologically conserved and were thought to be species poor until recent molecular data suggested otherwise [@moritzCryptic2018;@zozayaAre2019;@marinHidden2013]. Furthermore, shifts in diversification rates due to changes in Australian temperature and aridity have been reported in some squamate reptile clades including skinks [@raboskyExceptional2007] and geckos [@fujitaDiversification2010;@oliverContrasting2014;@oliverIndependent2014;@pepperGenetic2011;@pepperPalaeoclimate2011] as well in frogs [@catulloAridification2014;@vidal-garciaConvergent2015], saltbush [@kadereitMolecular2010], *Acacia* [@millerComparative2013], insects [@owenHow2017;@toussaintUnveiling2015], and plants [@crispHow2013]. Yet, the role of aridification and temperature changes on diversification is highly variable, with some arid-adapted vertebrates radiating following aridification [@raboskyExceptional2007;@pepperPalaeoclimate2011;@brennanMass2017] while other mesic-adapted groups have been driven extinct [@bryantLines2016;@cardilloPhylogeny2017;@owenHow2017]. Whether environmental change in Australia has had generally positive or negative influence on diversification rates in squamate reptiles is still unknown.

Today, the arid zone covers roughly 70% of the Australian landscape and provides habitat to 43% of all of Australia's terrestrial reptiles, termites, and ants [@byrneBirth2008]. Of these, roughly a quarter of the snake species diversity in Australia is represented by blindsnakes of the genus _Anilios_, which comprises **47** described species [@uetzQuarter2021] (Tiatragul et al., submitted MS). All currently described species of _Anilios_ are fossorial and are endemic to Australia except for _A. eryncinus_, which is endemic to New Guinea. blindsnakes are morphologically conserved, species can be difficult to distinguish, and previous phylogenetic and morphological studies have revealed that the true species diversity is underestimated [@marinHidden2013;@pyronSystematics2014;@hedgesTaxonomic2014;@ellisTyphlopid2019]. The diversification of _Anilios_ in Australia is reported to have originated from a single colonisation event by a mesic-adapted ancestor from Asia between 39 – 19 Ma [@vidalBlindsnake2010;@marinTracing2013]. Despite representing a substantial portion of snake biodiversity, we know little about the macroevolutionary dynamics of blindsnakes. The monophyly of the genus makes _Anilios_ well suited to test hypotheses about how lineages diversified with dramatic environmental changes during the Cenozoic because they are geographically restricted and can be found in every biome on mainland Australia.

@@

To our knowledge no study has assessed aridity as a continuous variable in an environment-dependent birth-death model framework, probably due to the difficulty of quantifying paleoaridity. However, recent developments in spatially-explicit paleoenvironmental reconstructions from the lithological and fossil record as well as global circulation models (GCMs) now enables us to estimate paleotemperature and paleoaridity data specific to Australia in the time span of interest. The use of these data sets allows us to conduct empirically analyses of the impacts of regionalised environmental changes on diversification rates over time for the first time.

In this study, we aimed to infer the historical biogeography of the Australian blindsnake radiation and investigate the impact of paleoenvironmental changes, specifically changes in temperature and aridity associated with Neogene aridification, on the diversification rates within this group. To do this, we fitted several birth-death models with reconstructed paleotemperature and aridity, and estimated diversification rates under these models. Additionally, we compared diversification rates between arid-adapted lineages versus mesic-adapted lineages to investigate if lineages that inhabit the arid biome have undergone faster or slower radiation.

@@

## Phylogenetic hypothesis and taxon sampling
<!-- Based on our phylogenetic tree -->
We employed the most comprehensive time-calibrated phylogeny of Australian blindsnakes with outgroups (Tiatragul et al., submitted MS). This tree includes 38 operational taxonomic units (OTUs) representing 35/47 described species of _Anilios_. The remaining 12 species could not be included because no molecular data or tissues samples are available. However, our phylogenetic data suggest the existence of at least three genetically distinct undescribed species, so we adjusted our taxon sampling fraction to 38/50 species. The genetically distinct OTUs that are not fully described species include three '_Anilios grypus_' (two in Western Australia and one in Queensland) and 2 '_Anilios ligatus_' (one from the eastern part of the range and one from the northern part) (Tiatragul et al., submitted MS). For ancestral biome estimation analyses, we also included _Ramphotyphlops multilineatus_, _Sundatyphlops polygrammicus_, and _Acutotyphlops subocularis_ as outgroup taxa.

@@

To investigate the biogeographic history of blindsnakes, we used the R package 'BioGeoBEARS' v.1.1.2 [@matzkeBioGeoBEARS2013]. BioGeoBEARS provides a flexible, likelihood-based framework that models how branches in the phylogeny (representing extant taxa and their ancestors) have shifted among a set of discrete biomes over time. These shifts in biomes can occur within a single branch (anagenetic events, including processes such as dispersal and extinction) or at the time of branching (cladogenetic events involving sympatric, vicariant, and founder-event speciation). 
We divided Australo-Papua into nine discrete biomes and assigned each Australian species to one or more biomes based on their occurrence records from the Atlas of Living Australia database as described in Tiatragul et al. (submitted). Biome classifications were based on the World Wildlife Fund's (WWF) Terrestrial Ecoregions of the World data set [@olsonTerrestrial2001] and followed the nomenclature of published Australia-specific studies [@cardilloPhylogeny2017;@pavon-vazquezComprehensive2021].

Using the phylogeny and distribution data, we fitted three widely used historical biogeographical methods, which vary in the kinds of cladogenetic and anagenetic events they model, including likelihood implementations of Dispersal-Vicariance Analysis (DIVALIKE) [@ronquistDispersalVicariance1997], Dispersal-Extinction-Cladogenesis (DEC) [@reeMaximum2008], and BayArea (BAYAREALIKE) [@landisBayesian2013] in BioGeoBEARS [@matzkeBioGeoBEARS2013]. We also fitted these three models, with two  additional relevant free parameters, both singly and together, for a total of 12 alternative models. The first additional parameter, j, assigns per-event weight to founder-events [@matzkeModel2014], and the second parameter, x, multiplies the probability of dispersal by distance to the power of x [@vandamEvaluating2016] (see Table S1 in Supporting Information). We used likelihood ratio tests to determine if the additional free parameter was a justified improvement over the simpler base models and selected the model that best fitted the data based on corrected Akaike Information Criterion (AICc) scores. Based on the best model, we ran an additional 50 biogeographic stochastic maps to account for uncertainty in ancestral biome estimation and transition between biomes, and to estimate the frequency and direction of dispersal events across biomes [@matzkeBioGeoBEARS2013].

@@

We used two alternative approaches to estimate paleoclimate in Australia from 40 Ma to the present-day. The first approach used information derived from paleo-Köppen climate bands from @scoteseAtlas2021. Köppen bands are estimated from different fossil and lithological indicators and give a broad categorisation of spatial variation in major climate types (tropical, sub-tropical arid, warm temperate, cool temperate, and polar). We obtained estimated Köppen climate bands at 5 Ma intervals from 40 Ma to 0 Ma from and used the sub-tropical arid Köppen band as a binary aridity index. We removed sites found in the ocean by using digital elevation models (DEMs) to mask sites below sea-level. To account for uncertainty in paleogeographic reconstructions of continental movements and sea-levels, we used two alterative DEMs [@straumeGlobal2020;@scotesePALEOMAP2018] to mask the ocean sites and provide two estimates of paleoaridity. To estimate paleotemperature from the Köppen bands, we followed the method of @scoteseAtlas2021 as implemented by @hagenEarth2021. Here, each Köppen climate band was assigned a temperature value based on present-day averages of those climate bands (tropical = 26, sub-tropical arid = 22, warm temperate = 16, cool temperate = 5, polar = -20). To account for global fluctuations in temperature, we then modified these values by a global temperature curve estimate from oxygen isotope data [@songSeawater2019]. Finally, to account for change of temperature with elevation, we used a lapse rate of 9.8°C/km. Following @hagenEarth2021, we model variation in the intensity of temperature oscillations during the Pleistocene. We also used the two alternative DEMs to give two alternative temperature curves using this approach [@straumeGlobal2020;@scotesePALEOMAP2018].

The second approach used a global circulation model (GCM) to obtain a model-based estimate of paleotemperature and paleoprecipitation [@valdesDeep2021]. The paleoclimatic data is based on a HadCM3L model which couples ocean, atmosphere and vegetation and simulated at 5 Ma intervals based on the paleogeography of @scoteseAtlas2021. To make it comparable with the binary aridity index derived from the Köppen band approach, we converted the paleoprecipitation into an aridity index by thresholding the precipitation data. We applied a threshold of 500mm/year, below which is considered arid, above which was considered non-arid. Each of the three global paleoaridity and paleotemperature maps, from the two approaches, were projected into a Behrmann equal-area projection using a spatial resolution of 110 km x 110 km. We then temporally interpolated the data from 5 Ma intervals to 100 ka using bilinear interpolation. Finally, we subset the data to include only the continent of Sahul from 40 Ma to the present-day. From these maps, we obtain multiple Sahulian curves for paleotemperature by taking the mean, minimum and maximum values within Sahul at each timestep. For paleoaridity, we took the mean value, as minimum and maximum showed little variation in the binary metric.

@@

## Data visualisation and software
ClaDS was implemented in Julia v.1.8.1 [@bezansonJulia2012], all other analyses were performed in R v.4.2.1 [@rcoreteamLanguage2022]. Figures were prepared using 'ggplot2' v.3.4.0 [@wickhamGgplot22016]. All map graphics were created using the Free and Open Source QGIS v.3.6. We arranged and finalised all figures using Adobe Illustrator v2022.

@@

<!-- ## values and info from 04_BioGeoBEARS_bsm_plots.R -->
Our BioGeoBEARS analyses all inferred tropical grassland as the ancestral biome of *Anilios*, but the ancestor leading to the crown *Anilios* is ambiguous (Fig. \@ref(fig:fig-biogeobears)a and Fig. S1). We found that the DEC+j+x model was the most preferred biogeographical history model (Δ AICc = -4.47, Table S2). A likelihood ratio test (LRT) showed that both founder events (+j parameter) and distance between biomes (+x parameter) together provide a statistically significant improvement over the simpler DEC model, DEC+x, and DEC+j (χ^2^, DF = 2, D-statistic = 26.7, p-value < 0.001).

Based on 50 BioGeoBEARS biogeographic stochastic mappings (BSM) of the DEC+j+x model, we estimated that an average of 71.2 ± 1.3 (mean ± SD) total anagenetic and cladogenetic events occurred throughout the phylogeny (see Table S3). Of these 40 ± 0 were cladogenetic events compared to 31.2 ± 1.3 anagenetic dispersal events. Out of the 40 cladogenetic dispersal events, most were narrow sympatry range expansion events (23.8 ± 2.1 events) followed by subset sympatry (5.2 ± 2.3 events), vicariance (4.9 ± 1.3), and founder events (6.1 ± 1.5). Our model also indicate that range expansion is expected to decrease as distance increase (free parameter x = -1.3).

The largest source of outward dispersal is the tropical grassland biome (16.2 ± 4.9 events), which makes up  43.6% compared outward dispersals from the arid biome (9.7 ± 3.2 events) of 26% (Fig. \@ref(fig:fig-biogeobears)b). Together the tropical grassland biome and the arid biome contributes the largest source of diversification accounting for 69.6% of outward transitions. On the other hand, the largest sinks for all inward dispersals are the tropical forest biome (7.9 ± 4.2 events; 21.2%) and the Mediterranean biome (7.0 ± 2.9 events; 18.8%). The greatest directional transition is from the arid biome to the Mediterranean biome (5.3 ± 1.5 events) and from the tropical grassland into the temperate forest biome (5.1 ± 1.3 events) (Fig. \@ref(fig:fig-biogeobears)c).

Based on current distributions from ALA, most _Anilios_ species are found entirely or at least partially in the tropical grassland biome (n = 20) followed by the arid biome (n = 13) and Mediterranean biome (n = 10). We identified divergent biogeographic transitions between two major groups of species, which we will refer to as Group 1 (n = 21 lineages, not including _A. proximus_) and Group 2 (n = 14 lineages, not including _A. silvia_ and _A. leucoproctus_) (Fig. \@ref(fig:fig-biogeobears)a). The common ancestor of Group 1 remained mostly in the tropical grassland biome, then 12/21 lineages expanded or transitioned into other biomes. The common ancestor of Group 2 first transitioned into the arid biome, and then 7/14 lineages expanded to the Mediterranean and temperate forest biomes, and/or transitioned back into the arid biome.

@@

```{r fig-RPANDA, fig.cap= 'Effect of aridity change on the diversification of Anilios estimated from the phylogeny. (a) The aridity index through time calculated from reconstructed paleoclimate (dotted line), and lineage through time estimated from the phylogeny (solid line). (b) Speciation rates through time from the best diversification model. (c) Association between speciation rate and aridity index. We found that increased aridity is associated with decrease speciation rate.'}

>>>>

**Running title** Decrease in blindsnake speciation correlate with aridification

@@

# Acknowledgements
ST would like to thank J Rolland and F. Condamine for helpful advice about fitting diversification models with RPANDA. Thanks to NJ Matzke and O Jiménez-Robles for help with BioGeoBEARS. Special thanks to IG Brennan, D Esquerré, M Pepper, CJ Pavón-Vázquez, and O Hagen for helpful advice on the analysis and discussion. JSK thanks the Australian Research Council for ongoing support.

@@

**Aim**: Shifts in diversification rates of Australian flora and fauna have been associated with aridification, but the relationship between diversification rates and aridity has never been quantified. We employed multiple approaches to accurately reconstruct paleoenvironments of Australia for the first time. We used this information, and phylogenetic-based analyses, to explore how changes in temperature and increasing aridity during the Neogene influenced the diversification of the Australian blindsnakes. We tested whether diversification rates differ between arid-adapted and mesic-adapted lineages.

**Taxon**: Typhlopidae, _Anilios_ blindsnakes

@@

**Results**: Ancestral area estimation indicated Australian blindsnakes have tropical grassland origins. We found that Australia-specific regional paleotemperature and paleoaridity provided a better explanation for diversification rate variation than global paleotemperature. Specifically, our best-fitting model indicated that speciation rates of blindsnakes decreased with increasing aridity. We found no difference in diversification rates between arid- and mesic-adapted lineages.

**Main conclusions**: Soon after dispersing to Australia, the common ancestors of Australian blindsnakes diversified rapidly in mesic habitats during the early Miocene. However, as the continent became increasingly arid, diversification rates decreased. We found that shifts in the environment led to the emergence of two major clades: one remaining in primarily mesic habitats and the other adapting to the expanding arid biome. Our results emphasise the importance of both arid and tropical biomes as sources and sinks of diversification.

@@

The present-day distribution of extant species can provide clues about their biogeographic history and insight into the tempo of their diversification. In particular, one major aim of evolutionary biology is to understand the role that environmental factors have played in shaping rates of species diversification [@simpsonTempo1944]. A variety of environmental correlates have been associated with diversification processes across clades, including elevation [@esquerreHow2019;@boschmanMountain2022], CO<sub>2</sub> concentration [@kergoatOpposite2018], microhabitats [@bars-closelDiversification2017], temperature [@ricklefsGlobal2006;@mayhewBiodiversity2012;@condamineMacroevolutionary2013;@condamineClimate2019;@rollandContribution2019;@sunRecent2020], and aridification [@raboskyExceptional2007;@pepperGenetic2011;@deepakAridification2018;@catulloAridification2014]. However, despite advancements in our understanding of the evolutionary relationships and divergence times of many clades, our knowledge of the impact of paleoenvironmental change on these clades remains limited, particularly in regard to the regional scale at which many clades radiate [@eyresClimatic2021]. As such, it remains a key challenge to combine regional paleoenvironmental histories, with molecular phylogenies, to understand the drivers of diversification over deep-time scales.

<!-- Terrestrial ectotherms, such as squamate reptiles, are highly sensitive to changes in external environment and climate. This can influence their diversification rates through multiple mechanisms. For instance, warmer temperature may increase resource availability, which in turn increases ecological niche space [@clarkeClimate2006]. Spatial and temporal heterogeneity in temperature can lead to climatic niche divergence, such as local adaptations in different geographic areas [@pepperGeodiversity2013], and changes in activity times of prey and predators [@santosTemperature2011;@lawsonLatitudinal2014]. Change in climate, leading to glaciation or aridification, has been shown to cause fragmentation of continuous habitats leading to isolated populations, which may lead to species-level divergence [@oliverRocks2019]. Ectotherms have higher metabolic rates in warm climates (i.e. Kleiber’s law), which may increase mutation rates and ultimately greater speciation rates [@allenKinetic2006;@makarievaMean2008]. Finally, aridification may offer new ecological niche for species to exploit, particularly mesic-adapted lineages that have colonised from elsewhere. Despite many possible pathways, little is known about whether environmental change is associated with variable rates of diversification. -->

The Australian continent has one of the most diverse reptile faunas on the planet and this may have been driven by its rapid environmental change in the Neogene period. After a long period of relatively stable climate and mesic habitats, the aridification of continental Australia began around 20 Ma (mega annum) with the development of pockets of aridity [@martinCenozoic2006;@byrneDecline2011]. Subsequent extreme warming and cooling periods in the last ~5 Ma creating what is now known as the "outback" [@byrneBirth2008;@pepperLife2021]. During this period, as aridification progressed, arid-adapted vegetation expanded while mesic biomes became increasingly fragmented [@crispHow2013;@martinCenozoic2006]. This shift in the environment also led to a radiation of some reptile clades, such as elapid snakes and skinks, which diversified into morphologically and ecologically distinct lineages [@leeDiversification2016;@sandersMolecular2008;@raboskyExceptional2007]. Conversely, other reptile clades, such as blindsnakes and geckos, remain morphologically conserved and were thought to be species poor until recent molecular data suggested otherwise [@moritzCryptic2018;@zozayaAre2019;@marinHidden2013]. Furthermore, shifts in diversification rates due to changes in Australian temperature and aridity have been reported in some squamate reptile clades including skinks [@raboskyExceptional2007] and geckos [@fujitaDiversification2010;@oliverContrasting2014;@oliverIndependent2014;@pepperGenetic2011;@pepperPalaeoclimate2011] as well in frogs [@catulloAridification2014;@vidal-garciaConvergent2015], insects [@owenHow2017;@toussaintUnveiling2015], and plants [@crispHow2013;@kadereitMolecular2010;@millerComparative2013]. Yet, the role of aridification and temperature changes on diversification is highly variable, with some arid-adapted vertebrates radiating following aridification [@raboskyExceptional2007;@pepperPalaeoclimate2011;@brennanMass2017] while other mesic-adapted groups have had suppressed diversification or been driven extinct [@bryantLines2016;@cardilloPhylogeny2017;@owenHow2017]. Whether environmental change in Australia has had generally positive or negative influence on diversification rates in squamate reptiles is still unknown.

Today, the arid zone covers roughly 70% of the Australian landscape and provides habitat to 43% of all of Australia’s terrestrial reptiles [@byrneBirth2008].Of these, roughly a quarter of the snake species diversity in Australia is represented by blindsnakes of the genus *Anilios*, which currently comprises 47 described species  **47** described species [@uetzQuarter2021] (Tiatragul et al., submitted MS). All but three species of currenty described _Anilios_ are endemic to Australia, _A. torresianus_ and _A. leucoproctus_ also occur in southern parts of New Guinea, while _A. eryncinus_ has only been recorded in New Guinea and the Maluku Islands. 
Blindsnakes are morphologically conserved and previous phylogenetic and morphological studies have revealed that the true species diversity is underestimated [@marinHidden2013;@pyronSystematics2014;@hedgesTaxonomic2014;@ellisTyphlopid2019]. The diversification of _Anilios_ in Australia is reported to have originated from a single dispersal event by a mesic-adapted ancestor from Southeast Asia or Indonesia between 39 – 19 Ma [@vidalBlindsnake2010;@marinTracing2013]. 
However, phylogenetic analyses from Tiatragul et al. (2023) indicated that the closest relative to all *Anilios* is *Ramphotyphlops multilineatus*, a species that occur in Western New Guinea, Kai Islands, and Salawati, suggesting the Australian radiation may have a Melanesian origin
Despite representing a substantial portion of snake diversity, we know little about the macroevolutionary dynamics of blindsnakes. The monophyly of the genus makes _Anilios_ well suited to test hypotheses about how lineages diversified with dramatic environmental changes during the Cenozoic because they are geographically restricted and can be found in every biome on mainland Australia.

@@

To our knowledge no study has assessed aridity as a continuous variable in an environment-dependent birth-death model framework, probably due to the difficulty of quantifying paleoaridity. However, recent developments in spatially-explicit paleoenvironmental reconstructions from the lithological and fossil record as well as global circulation models (GCMs) now enables us to estimate paleotemperature and paleoaridity data specific to Australia in the time span of interest. The use of these data sets allows us to conduct empirical analyses of the impacts of regionalised environmental changes on diversification rates over time for the first time.

In this study, we aimed to infer the historical biogeography of the Australian blindsnake radiation and investigate the impact of paleoenvironmental changes, specifically changes in temperature and aridity associated with Neogene aridification, on diversification rates within this group. To do this, we fitted several birth-death models with reconstructed paleotemperature and aridity, and estimated diversification rates under these models. Additionally, we compared diversification rates between arid-adapted lineages versus mesic-adapted lineages to investigate if lineages that inhabit the arid biome have undergone faster or slower radiation.

@@

## Phylogenetic hypothesis and taxon sampling

We used the most updated phylogeny for *Anilios* and closely related outgroups from Tiatragul et al., submitted MS. The species tree was built with the shortcut coalescent method in ASTRAL III [@zhangASTRALIIIPolynomialTime2018] using maximum likelihood gene trees inferred in IQTREE from 4930 loci (322 Anchored Hybrid Enrichment, 4574 ultraconserved elements, and 34 legacy genes) for 35/47 recognised *Anilios* species. Tiatragul et al. (submitted) time-calibrated the phylogeny using 27 outgroup squamate fossils as primary calibrations using MCMCTree from the PAML package [@yangPAML2007] (see Tiatragul et al., submitted MS for details). Phylogenetic data in Tiatragul et al. (submitted) suggest the existence of at least three genetically distinct undescribed species, so we followed their designation and adjusted our taxon sampling fraction to 38/50 species. 
The genetically distinct OTUs that are not fully described species include three '_Anilios grypus_' (two in Western Australia and one in Queensland) and two '_Anilios ligatus_' (one from the eastern part of the range and one from the northern part) (Tiatragul et al., submitted MS). 
For ancestral biome estimation analyses, we included three outgroup blindsnakes species from Southeast Asia, Indonesia, and Melanesia, where the ancestor of *Anilios* has been hypothesised to have originated [@vidalBlindsnake2010]. The three species include *Ramphotyphlops multilineatus* (Western New Guinea, Kai Island, and Salawati), *Sundatyphlops polygrammicus* (Moyo Island, Indonesia), and *Acutotyphlops subocularis* (Bismarck Archipelago). We were not able to include a *Ramphotyphlops* from continental Southeast Asia in the time-calibrated phylogeny because insufficient data was available for divergence time estimation.

@@

To investigate the biogeographic history of blindsnakes, we used the R package 'BioGeoBEARS' v.1.1.2 [@matzkeBioGeoBEARS2013]. BioGeoBEARS provides a flexible, likelihood-based framework that models how branches in the phylogeny (representing extant taxa and their ancestors) have shifted among a set of discrete biomes over time. These shifts in biomes can occur within a single branch (anagenetic events, including processes such as dispersal and extinction) or at the time of branching (cladogenetic events involving sympatric, vicariant, and founder-event speciation). 
We divided Australo-Papua into nine discrete biomes and assigned each Australian species to one or more biomes based on their occurrence records from the Atlas of Living Australia database. Biome classifications were based on the World Wildlife Fund's (WWF) Terrestrial Ecoregions of the World data set [@olsonTerrestrial2001] and followed the nomenclature of published Australia-specific studies [@cardilloPhylogeny2017;@pavon-vazquezComprehensive2021].

Using the phylogeny and distribution data, we fitted models that vary in the kinds of cladogenetic and anagenetic events they model, including likelihood implementations of Dispersal-Vicariance Analysis (DIVALIKE) [@ronquistDispersalVicariance1997], Dispersal-Extinction-Cladogenesis (DEC) [@reeMaximum2008], and BayArea (BAYAREALIKE) [@landisBayesian2013] in BioGeoBEARS [@matzkeBioGeoBEARS2013]. We also fitted these three models, with two  additional relevant free parameters, both singly and together, for a total of 12 alternative models. The first additional parameter, j, assigns per-event weight to founder-events [@matzkeModel2014], and the second parameter, x, multiplies the probability of dispersal by distance to the power of x [@vandamEvaluating2016] (Table S1). We used likelihood ratio tests to determine if the additional free parameter was a justified improvement over the simpler base models and selected the model that best fitted the data based on corrected Akaike Information Criterion (AICc) scores. Based on the best model, we ran an additional 50 biogeographic stochastic maps to account for uncertainty in ancestral biome estimation and transition between biomes, and to estimate the frequency and direction of dispersal events across biomes [@matzkeBioGeoBEARS2013].

@@

We used two alternative approaches to estimate paleoclimate in Australia from 40 Ma to the present-day. The first approach used information derived from paleo-Köppen climate bands from @scoteseAtlas2021. Köppen bands are estimated from different fossil and lithological indicators and give a broad categorisation of spatial variation in major climate types (tropical, sub-tropical arid, warm temperate, cool temperate, and polar). We obtained estimated Köppen climate bands at 5 Ma intervals from 40 Ma to 0 Ma and used the sub-tropical arid Köppen band as a binary aridity index. We removed sites found in the ocean by using digital elevation models (DEMs) to mask sites below sea-level. To account for uncertainty in paleogeographic reconstructions of continental movements and sea-levels, we used two alternative DEMs [@straumeGlobal2020;@scotesePALEOMAP2018] to mask the ocean sites and provide two estimates of paleoaridity. To estimate paleotemperature from the Köppen bands, we followed the method of @scoteseAtlas2021 as implemented by @hagenEarth2021. Here, each Köppen climate band was assigned a temperature value based on present-day averages of those climate bands (tropical = 26, sub-tropical arid = 22, warm temperate = 16, cool temperate = 5, polar = -20). To account for global fluctuations in temperature, we then modified these values by a global temperature curve estimate from oxygen isotope data [@songSeawater2019]. Finally, to account for change of temperature with elevation, we used a lapse rate of 9.8°C/km. Following @hagenEarth2021, we model variation in the intensity of temperature oscillations during the Pleistocene. We also used the two alternative DEMs to give two alternative temperature curves using this approach [@straumeGlobal2020;@scotesePALEOMAP2018].

The second approach used a global circulation model (GCM) to obtain a model-based estimate of paleotemperature and paleoprecipitation [@valdesDeep2021]. The paleoclimatic data is based on a HadCM3L model which couples ocean, atmosphere and vegetation and simulated at 5 Ma intervals based on the paleogeography of @scoteseAtlas2021. To make it comparable with the binary aridity index derived from the Köppen band approach, we converted the paleoprecipitation into an aridity index by thresholding the precipitation data. We applied a threshold of 500mm/year, below which is considered arid, above which was considered non-arid [@holzapfelDeserts2008;@waltonArid1969]. Each of the three global paleoaridity and paleotemperature maps, from the two approaches, were projected into a Behrmann equal-area projection using a spatial resolution of 110 km x 110 km. We then temporally interpolated the data from 5 Ma intervals to 100 ka using bilinear interpolation. Finally, we subset the data to include only the continent of Sahul from 40 Ma to the present-day. From these maps, we obtain multiple Sahulian curves for paleotemperature by taking the mean, minimum and maximum values within Sahul at each timestep. For paleoaridity, we took the mean value, as minimum and maximum showed little variation in the binary metric.

@@

PANDA was implemented in Julia v.1.8.1 [@bezansonJulia2012], all other analyses were performed in R v.4.2.1 [@rcoreteamLanguage2022]. Figures were prepared using 'ggplot2' v.3.4.0 [@wickhamGgplot22016]. All map graphics were created using the Free and Open Source QGIS v.3.6. We arranged and finalised all figures using Adobe Illustrator v2022.

@@

<!-- ## values and info from 04_BioGeoBEARS_bsm_plots.R -->
Our BioGeoBEARS analyses all inferred tropical grassland as the ancestral biome of *Anilios*, but the ancestral biome for the common ancestor between *Ramphotyphlops multilineatus* and *Anilios* is ambiguous (Fig. \@ref(fig:fig-biogeobears)a and Fig. S1). We found that the DEC+j+x model was the best-fitting biogeographical history model (Δ AICc = -4.47, Table S2). A likelihood ratio test (LRT) showed that both founder events (+j parameter) and distance between biomes (+x parameter) together provide statistically significant improvements over the simpler DEC, DEC+x, and DEC+j models (χ^2^, DF = 2, D-statistic = 26.7, p-value < 0.001).

Based on 50 BioGeoBEARS biogeographic stochastic mappings (BSM) of the DEC+j+x model, we estimated that an average of 71.2 ± 1.3 (mean ± SD) total biome shifts occurred throughout the phylogeny (see Table S3). Of these 40 ± 0 were cladogenetic (branching) events compared to 31.2 ± 1.3 anagenetic (within a single branch) dispersal events. Out of the 40 branching events, in 23.8 ± 2.1 events sister species maintained their ancestral biome (“narrow sympatry”) and in 5.2 ± 2.3 events one sister species maintain their ancestral biome while the other sister occupies only a subset of the ancestral range (“subset sympatry”). The remaining branching events were made up by 6.1 ± 1.5 founder events where one of the sister species occupies a new range, and 4.9 ± 1.3 vicariance events where the sister species split their ancestral range. Our model also indicate that range expansion is expected to decrease as distance increase (free parameter x = -1.3).

Based on current distributions from ALA, most Anilios species are found entirely or partially in the tropical grassland biome (n = 20) followed by the arid biome (n = 13) and Mediterranean biome (n = 10). Apart from A. silvia, A. leucoproctus, and A. proximus that appeared to form independent divergent lineages (Fig. \@ref(fig:fig-biogeobears)a), we identified divergent biogeographic transitions between two major groups of species, which we refer to as Group 1 (n = 21 lineages) and Group 2 (n = 14 lineages). The common ancestor of Group 1 remained mostly in the tropical grassland biome, then 12/21 lineages expanded or transitioned into other biomes. The common ancestor of Group 2 first transitioned into the arid biome, four lineages remained in the arid biome while ten lineages transitioned or expanded into other biomes.

The tropical grassland biome is the largest source of outward dispersal (16.2 ± 4.9 events; 46.6%) followed by the arid biome (9.7 ± 3.2 events; 26%)(Fig. 1b). On the other hand, the largest sinks for all inward dispersals are the tropical forest biome (7.9 ± 4.2 events; 21.2%) and the Mediterranean biome (7.0 ± 2.9 events; 18.8%). The greatest directional transition is from the arid biome to the Mediterranean biome (5.3 ± 1.5 events) and from the tropical grassland biome into the temperate forest biome (5.1 ± 1.3 events) (Fig. \@ref(fig:fig-biogeobears)c).

@@

```{r fig-RPANDA, fig.cap= 'Effect of aridity change on the diversification of Anilios estimated from the phylogeny. (a) The aridity index through time calculated from reconstructed paleoclimate (dotted line), and lineage through time estimated from the phylogeny (solid line). (b) Fitted diversification rates through time reconstructed from parameter estimates from the best diversification model (speciation rates vary exponentially with paleoaridity and constant extinction rate; “BEnvVar + DCST (EXPO)”). (c) Fitted diversification rates at different aridity index values calculated from parameter estimates of the previous model. Increased aridity is associated with decrease net diversification rates in this model..'}
